---
title: "Homework 4"
author: "Jake Perez"
format: gfm
editor: visual
---

## Using ggplot to visualize data

The `tidyverse` package includes functions that are great for data visualization, such as `ggplot`.

```{r, message = FALSE}
library(tidyverse)
```

```{r}
my_data <- read_csv('iron_raw_data_04.csv')
```

We can now use ggplot to make a custom figure. We can use the following code block to create a violin plot of sediment layer color with the mass of the sample used for analyses.

```{r}
ggplot(my_data) +
  geom_violin(aes(x = color, y = sample_mass), trim = FALSE) + 
  geom_boxplot(aes(x = color, y = sample_mass), width = 0.1) + 
  xlab("Layer Color") + ylab("Sample Mass (g)")
```

The violin plot is envoked with the `geom_violin()` function and writing `trim = FALSE` puts the tails on the ends of the plot. A warning is given due to the Light Grey Category having only one sample represented in the data set and therefore `ggplot` does not produce a violin graph.

We can also layer additional graphs to the figure, such as a box-and-whisker plot to show additional statistical data like the sample medians and quartiles. This is done using the `geom_boxplot()` function. We specify the `width` parameter here to make the boxplot fit within the violin plot.

To make the axes more descriptive than the data frame column headers, we can write the labels using the `xlab()` and `ylab()` functions.
