---
title: "Homework 4"
author: "Jake Perez"
format: gfm
editor: visual
---

## Using ggplot to visualize data

The `tidyverse` package includes functions that are great for data visualization, such as `ggplot`.

```{r, message = FALSE}
library(tidyverse)
my_data <- read_csv('iron_raw_data_04.csv')
```

We can now use ggplot to make a custom figure. We can use the following code block to create a violin plot of sediment layer color with the mass of the sample used for analyses.

```{r}
#Create a vector for the color palette of the violin plot
color.palette <- c("gray10", "forestgreen", "gray50", "white")

#Create an object for the violin plot (with various customizations)
p <- ggplot(my_data, aes(x = color, y = sample_mass, fill = color)) +
  geom_violin(color = "black", alpha = 0.65, trim = FALSE) + 
  scale_fill_manual(values = color.palette)

#Layer a boxplot onto the violin plot
p + geom_boxplot(aes(x = color, y = sample_mass), fill = "white", width = 0.1) +
  theme(legend.position = "none") + theme_classic() + 
  xlab("Layer Color") + ylab("Sample Mass (g)")
```

The violin plot is envoked with the `geom_violin()` function and writing `trim = FALSE` puts the tails on the ends of the plot. A warning is given due to the Light Grey Category having only one sample represented in the data set and therefore `ggplot` does not produce a violin graph.

We can also layer additional graphs to the figure, such as a box-and-whisker plot to show additional statistical data like the sample medians and quartiles. This is done using the `geom_boxplot()` function. We specify the `width` parameter here to make the boxplot fit within the violin plot.

To make the axes more descriptive than the data frame column headers, we can write the labels using the `xlab()` and `ylab()` functions.

Since the x-axis is a color category, it would be most intuitive to color the violin plots according to the color of the layer. The `color.palette` object establishes the color codes I decided to use for the violin plots (note: there are only four values specified since the Light Grey category is not represented in the violin plots). You can also use `alpha` to adjust the transparency of the object. This helps the overlying box plots become more visible.

This is a good representation of the following data because it shows how sample mass was distributed for each sediment layer color type because the geochemical analyses that follow afterwards may be impacted by how large the sample mass was. Therefore, it is a preliminary step to determine if accounting for that impact may non-uniformly affect our color categories.
